Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-25 app-datastruct.tex
Use breakurl package to prevent the bitbucket url from overflowing into the margin.
Fixed inconsistencies with reference to the ZEUS hydro method.  In some places it was 'Zeus', in others 'ZEUS'.  Now all calls are to the \zeus command (defined in local-commands.tex), and can be adjusted however the heck one wants.

\subsection{Outline and List of Routines.}
\begin{enumerate}
 \item Creation and Destruction
   \begin{enumerate}
     \item Flow of Rebuild hierarchy: 
       \begin{enumerate}
	 \item delete hierarchy
	 \item build new hierarchy
	 \item scavange from old hierarchy
       \end{enumerate}
     \item Cell Flagging Methods
     \item Grid Fracturing Methods
   \end{enumerate}    
 \item Parallelism
  \begin{enumerate}
    \item Top Grid (block domain decomposition)
    \item Subgrids (memory balance, level by level)(Actually, I think
          its Parent Grid by Parent Grid.)
  \end{enumerate}
 \item Relation to other objects: 
   \begin{enumerate}
     \item Parents (No refinement jump requirements.  Talk about flux
        correction improvement to this effect here?)
     \item Siblings
     \item Children
     \end{enumerate}
 \item Traversing the hierarchy (this might be rolled into the
   previous bullet)
  \begin{enumerate}
    \item Level iterator 
    \item Hierarchy iterator
    \item Grid Array
    \item Chaining Mesh
  \end{enumerate}
  \item Particle treatement
   \begin{enumerate}
     \item Which grid gets the particles
     \item How particles are refined: criteria
     \item How particles are refined: how it fits in with the Grid refinement
   \end{enumerate}
  \item Ghost zones
    \begin{enumerate}
      \item how, when filled
    \end{enumerate}
  \item Timestepping
    \begin{enumerate}
      \item How timestep is determined
      \item Order of level operations: W cycle
    \end{enumerate}
\end{enumerate}
A list of routines to mention: \red{Zeus does this in a big table that
  references code name, what it does, and what equations it
  solves. Might be a useful idea to steal.}
\begin{description}
 \item[\tt{EvolveHierarchy}] Primary timestep control routine. 
 \item[\tt{EvolveLevel}] Primary evolution routine.
 \item[\tt{RebuildHierarchy}] Rebuilds the grid hierarchy
\end{description}

@@

A {\tt ProtoSubgrid} that fails either the efficency or size
requirement is then broken into smaller {\tt ProtoSubgrid}s by a
method that tries to preserve physically continuous structures onto
individual grids.  The process examines the 1d Signature of the Flagging
field, which is the projection of the Flagging Field along two of the
coordinate axes.  First, each of the 3 signature fields is examined
for zeros.  Zeros in the signature field indicates disjoint
structures, so the {\tt ProtoSubgrid} is first cut into two smaller
ones along that line.  If no such cuts can be made, zero crossings of
the second derivative in the signature are then looked for.  Zero
crossings in the second derivative indicates edges of structures.  If
more than one exists, the 'strongest crossing,' that is the largest
jump, is selected for the cut.  If no such zero crossings can be
found, the {\tt ProtoSubgrid} is then cut evenly along the longest
axis.  See figure \ref{fig.signature} for a schematic of this process.

>>>>

\subsection{Outline and List of Routines.}
\begin{enumerate}
 \item Creation and Destruction
   \begin{enumerate}
     \item Flow of Rebuild hierarchy: 
       \begin{enumerate}
	 \item delete hierarchy
	 \item build new hierarchy
	 \item scavange from old hierarchy
       \end{enumerate}
     \item Cell Flagging Methods
     \item Grid Fracturing Methods
   \end{enumerate}    
 \item Parallelism
  \begin{enumerate}
    \item Top Grid (block domain decomposition)
    \item Subgrids (memory balance, level by level)(Actually, I think
          its Parent Grid by Parent Grid.)
  \end{enumerate}
 \item Relation to other objects: 
   \begin{enumerate}
     \item Parents (No refinement jump requirements.  Talk about flux
        correction improvement to this effect here?)
     \item Siblings
     \item Children
     \end{enumerate}
 \item Traversing the hierarchy (this might be rolled into the
   previous bullet)
  \begin{enumerate}
    \item Level iterator 
    \item Hierarchy iterator
    \item Grid Array
    \item Chaining Mesh
  \end{enumerate}
  \item Particle treatement
   \begin{enumerate}
     \item Which grid gets the particles
     \item How particles are refined: criteria
     \item How particles are refined: how it fits in with the Grid refinement
   \end{enumerate}
  \item Ghost zones
    \begin{enumerate}
      \item how, when filled
    \end{enumerate}
  \item Timestepping
    \begin{enumerate}
      \item How timestep is determined
      \item Order of level operations: W cycle
    \end{enumerate}
\end{enumerate}
A list of routines to mention: \red{\zeus\ does this in a big table that
  references code name, what it does, and what equations it
  solves. Might be a useful idea to steal.}
\begin{description}
 \item[\tt{EvolveHierarchy}] Primary timestep control routine. 
 \item[\tt{EvolveLevel}] Primary evolution routine.
 \item[\tt{RebuildHierarchy}] Rebuilds the grid hierarchy
\end{description}

@@

A {\tt ProtoSubgrid} that fails either the efficency or size
requirement is then broken into smaller {\tt ProtoSubgrid}s by a
method that tries to preserve physically continuous structures onto
individual grids.  The process examines the 1d Signature of the Flagging
field, which is the projection of the Flagging Field along two of the
coordinate axes.  First, each of the 3 signature fields is examined
for zeros.  Zeros in the signature field indicates disjoint
structures, so the {\tt ProtoSubgrid} is first cut into two smaller
ones along that line.  If no such cuts can be made, zero crossings of
the second derivative in the signature are then looked for.  Zero
crossings in the second derivative indicates edges of structures.  If
more than one exists, the ``strongest crossing'', that is the largest
jump, is selected for the cut.  If no such zero crossings can be
found, the {\tt ProtoSubgrid} is then cut evenly along the longest
axis.  See figure \ref{fig.signature} for a schematic of this process.

### 2013-06-25 numerical-amr.tex
Fixed inconsistencies with reference to the ZEUS hydro method.  In some places it was 'Zeus', in others 'ZEUS'.  Now all calls are to the \zeus command (defined in local-commands.tex), and can be adjusted however the heck one wants.

Each grid patch in \enzo\ contains arrays of values for baryon and particle quantities.   Grids are partitioned into a core of \emph{active zones} and a surrounding layer of \emph{ghost zones}, as shown in Figure~\ref{fig.amr_hierarchy}.  The active zones store field values and ghost zones are used to temporarily store values that have been obtained directly from neighboring grids or interpolated from a parent grid.  These zones are necessary to accommodate the computational stencil of the (magneto)hydrodynamics solvers (Sections~\ref{sec.hydro.ppm} through~\ref{sec.hydro.zeus}) and the gravity solver (Section~\ref{sec.gravity}).  The PPM and Zeus hydro solvers require 3 layers of ghost zones,  and the gravity solver requires 6 (although only for the density and potential fields).  This can lead to significant memory and computational overhead, particularly for smaller grid patches at high levels of refinement.

@@

%Each grid patch in \enzo\ contains arrays of values for baryon and 
%particle quantities.   Grids are 
%partitioned into a core of \emph{real zones} and a surrounding layer 
%of \emph{ghost zones}.  The real zones store field values and ghost 
%zones are used to temporarily store values which have been obtained 
%directly from neighboring grids or interpolated from a parent grid.  
%These zones are necessary to accommodate the computational stencil of 
%the hydrodynamics solvers (Sections~\ref{sec.ov.hydro.ppm} and~\ref{sec.ov.hydro.zeus}) 
%and the gravity solver (Section~\ref{sec.ov.nbody}).  
%The PPM and Zeus hydro solvers require 3 layers of ghost zones,  and the gravity
%solver requires 6.  This can lead to significant memory 
%and computational overhead, particularly for smaller grid patches
%at high levels of refinement.  More information on filling of ghost
%zones can be found in the appendix.

>>>>

Each grid patch in \enzo\ contains arrays of values for baryon and particle quantities.   Grids are partitioned into a core of \emph{active zones} and a surrounding layer of \emph{ghost zones}, as shown in Figure~\ref{fig.amr_hierarchy}.  The active zones store field values and ghost zones are used to temporarily store values that have been obtained directly from neighboring grids or interpolated from a parent grid.  These zones are necessary to accommodate the computational stencil of the (magneto)hydrodynamics solvers (Sections~\ref{sec.hydro.ppm} through~\ref{sec.hydro.zeus}) and the gravity solver (Section~\ref{sec.gravity}).  The PPM and \zeus\ hydro solvers require 3 layers of ghost zones,  and the gravity solver requires 6 (although only for the density and potential fields).  This can lead to significant memory and computational overhead, particularly for smaller grid patches at high levels of refinement.

@@

%Each grid patch in \enzo\ contains arrays of values for baryon and 
%particle quantities.   Grids are 
%partitioned into a core of \emph{real zones} and a surrounding layer 
%of \emph{ghost zones}.  The real zones store field values and ghost 
%zones are used to temporarily store values which have been obtained 
%directly from neighboring grids or interpolated from a parent grid.  
%These zones are necessary to accommodate the computational stencil of 
%the hydrodynamics solvers (Sections~\ref{sec.ov.hydro.ppm} and~\ref{sec.ov.hydro.zeus}) 
%and the gravity solver (Section~\ref{sec.ov.nbody}).  
%The PPM and \zeus\ hydro solvers require 3 layers of ghost zones,  and the gravity
%solver requires 6.  This can lead to significant memory 
%and computational overhead, particularly for smaller grid patches
%at high levels of refinement.  More information on filling of ghost
%zones can be found in the appendix.

### 2013-06-25 numerical-gravity.tex
Fixed inconsistencies with reference to the ZEUS hydro method.  In some places it was 'Zeus', in others 'ZEUS'.  Now all calls are to the \zeus command (defined in local-commands.tex), and can be adjusted however the heck one wants.

For isolated boundary conditions, we use the James method \citep{James77}.  In this case, the Greens function is generated in real-space so as to have the correct zero-padding properties and then transformed into the Fourier domain.  In both cases, the potential is then transformed back into the real domain to get potential values at the cell centers.  These are differenced with a two-point centered difference scheme to obtain accelerations at the cell centers (except if we are using the staggered Zeus-like solver, in which case the accelerations are computed at the cell faces to match the velocities).  Particle accelerations are obtained using a (linear) CIC interpolation from the grid.

>>>>

For isolated boundary conditions, we use the James method \citep{James77}.  In this case, the Greens function is generated in real-space so as to have the correct zero-padding properties and then transformed into the Fourier domain.  In both cases, the potential is then transformed back into the real domain to get potential values at the cell centers.  These are differenced with a two-point centered difference scheme to obtain accelerations at the cell centers (except if we are using the staggered \zeus-like solver, in which case the accelerations are computed at the cell faces to match the velocities).  Particle accelerations are obtained using a (linear) CIC interpolation from the grid.

### 2013-06-25 numerical-overview.tex
Fixed inconsistencies with reference to the ZEUS hydro method.  In some places it was 'Zeus', in others 'ZEUS'.  Now all calls are to the \zeus command (defined in local-commands.tex), and can be adjusted however the heck one wants.

\subsubsection{The Zeus hydro method}\label{sec.ov.hydro.zeus}

@@

Each grid patch in \enzo\ contains arrays of values for baryon and 
particle quantities.   Grids are 
partitioned into a core of \emph{real zones} and a surrounding layer 
of \emph{ghost zones}.  The real zones store field values and ghost 
zones are used to temporarily store values which have been obtained 
directly from neighboring grids or interpolated from a parent grid.  
These zones are necessary to accommodate the computational stencil of 
the hydrodynamics solvers (Sections~\ref{sec.ov.hydro.ppm} and~\ref{sec.ov.hydro.zeus}) 
and the gravity solver (Section~\ref{sec.ov.nbody}).  
The PPM and Zeus hydro solvers require 3 layers of ghost zones,  and the gravity
solver requires 6.  This can lead to significant memory 
and computational overhead, particularly for smaller grid patches
at high levels of refinement.  More information on filling of ghost
zones can be found in the appendix.

>>>>

\subsubsection{The \zeus\ hydro method}\label{sec.ov.hydro.zeus}

@@

Each grid patch in \enzo\ contains arrays of values for baryon and 
particle quantities.   Grids are 
partitioned into a core of \emph{real zones} and a surrounding layer 
of \emph{ghost zones}.  The real zones store field values and ghost 
zones are used to temporarily store values which have been obtained 
directly from neighboring grids or interpolated from a parent grid.  
These zones are necessary to accommodate the computational stencil of 
the hydrodynamics solvers (Sections~\ref{sec.ov.hydro.ppm} and~\ref{sec.ov.hydro.zeus}) 
and the gravity solver (Section~\ref{sec.ov.nbody}).  
The PPM and \zeus\ hydro solvers require 3 layers of ghost zones,  and the gravity
solver requires 6.  This can lead to significant memory 
and computational overhead, particularly for smaller grid patches
at high levels of refinement.  More information on filling of ghost
zones can be found in the appendix.

### 2013-06-25 table_of_variables_and_constants.tex
Fixed inconsistencies with reference to the ZEUS hydro method.  In some places it was 'Zeus', in others 'ZEUS'.  Now all calls are to the \zeus command (defined in local-commands.tex), and can be adjusted however the heck one wants.

  \startdata
  \multicolumn{2}{l}{\underline{Fluid Quantities}} \\[5pt]
  $E$ & comoving total fluid energy density \\
  $e$ & comoving thermal energy density \\
  $T$ & baryon temperature \\
  $S$ & fluid entropy \\
  $\rho$ & comoving baryonic density \\
  $\rho_{\!X}$ & mass density of component $X$ (gas, dark matter, stars, metals, total, etc.) \\
  $t$ & time \\
  $\vecv$ & peculiar velocity \\
  $\vecB$ & comoving magnetic field strength \\
  $p$ & thermal pressure \\
  $p^*$ & total (thermal + magnetic) isotropic pressure \\
  $\gamma$ & ideal gas ratio of specific heats \\
  $\phi$ & gravitational potential \\
  $\Lambda$ & radiative/chemical cooling rate \\
  $\Gamma$ & radiative heating rate \\
  $Q$ & artificial viscosity stress tensor (only for ZEUS hydro) \\[3pt]

>>>>

  \startdata
  \multicolumn{2}{l}{\underline{Fluid Quantities}} \\[5pt]
  $E$ & comoving total fluid energy density \\
  $e$ & comoving thermal energy density \\
  $T$ & baryon temperature \\
  $S$ & fluid entropy \\
  $\rho$ & comoving baryonic density \\
  $\rho_{\!X}$ & comoving mass density of component $X$ (gas, dark matter, stars, metals, total, etc.) \\
  $t$ & time \\
  $\vecv$ & peculiar velocity \\
  $\vecB$ & comoving magnetic field strength \\
  $p$ & thermal pressure \\
  $p^*$ & total (thermal + magnetic) isotropic pressure \\
  $\gamma$ & ideal gas ratio of specific heats \\
  $\phi$ & gravitational potential \\
  $\Lambda$ & radiative/chemical cooling rate \\
  $\Gamma$ & radiative heating rate \\
  $Q$ & artificial viscosity stress tensor (only for \zeus\ hydro) \\[3pt]

### 2013-06-25 tests-shockpool.tex
Fixed inconsistencies with reference to the ZEUS hydro method.  In some places it was 'Zeus', in others 'ZEUS'.  Now all calls are to the \zeus command (defined in local-commands.tex), and can be adjusted however the heck one wants.

For Zeus (middle row of Figure~\ref{fig.shockpool}), the shock is broadened because of artificial viscosity and there are slightly more post-shock oscillations, although again quite mild.  The impact of entering and exiting the refined region is somewhat larger than in PPM; however, the most noticeable difference is the incorrect position of the shock front. This is due to the fact that the scheme is not energy-conserving (see also the Sedov problem in Section~\ref{sec.tests.sedov}), and has very little to do with the passage through the refined region.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/ShockPool}
\caption{This plot shows, for each column, three snapshots at $t=0.1, 0.15$ and 0.35 of a $\mathcal{M}=2$ shock as it propagates through the grid with a static refined region extends from $x = 0.25$ to 0.75 (shown in grey in each panel).   Each row shows the result for a different solver (PPM/Zeus/MUSCL from top to bottom).  The solid line shows the analytic solution.  Note that we focus each panel on a small region of the entire domain to better show the shock front.}
\label{fig.shockpool}
\end{center}
\end{figure}

>>>>

For \zeus\ (middle row of Figure~\ref{fig.shockpool}), the shock is broadened because of artificial viscosity and there are slightly more post-shock oscillations, although again quite mild.  The impact of entering and exiting the refined region is somewhat larger than in PPM; however, the most noticeable difference is the incorrect position of the shock front. This is due to the fact that the scheme is not energy-conserving (see also the Sedov problem in Section~\ref{sec.tests.sedov}), and has very little to do with the passage through the refined region.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/ShockPool}
\caption{This plot shows, for each column, three snapshots at $t=0.1, 0.15$ and 0.35 of a $\mathcal{M}=2$ shock as it propagates through the grid with a static refined region extends from $x = 0.25$ to 0.75 (shown in grey in each panel).   Each row shows the result for a different solver (PPM/\zeus/MUSCL from top to bottom).  The solid line shows the analytic solution.  Note that we focus each panel on a small region of the entire domain to better show the shock front.}
\label{fig.shockpool}
\end{center}
\end{figure}

### 2013-06-25 tests-wavepool.tex
Fixed inconsistencies with reference to the ZEUS hydro method.  In some places it was 'Zeus', in others 'ZEUS'.  Now all calls are to the \zeus command (defined in local-commands.tex), and can be adjusted however the heck one wants.

Figure~\ref{fig.wavepool} shows the evolution of the wave at three different times ($t = 0.2, 0.3$ and 0.8) for three of our solvers (PPM, Zeus, and MUSCL).  In all panels, the shaded regions denote the statically refined region. The leftmost column shows the wave just before it enters the refined region so that we can gauge how the solver is operating in the absence of AMR.  The center column shows the wave after it has fully entered the refined region.  The rightmost column shows the wave well after it has exited the refined region.

@@

For the Zeus solver, we see that even before it has entered the refined region there are small oscillations excited behind the wave, although it is also worth noting that the wave itself is beautifully propagated without significant damping or phase errors.  In this test we use only our standard, low amount of quadratic artificial viscosity.  These waves could be damped by additional viscosity, but we do not add any in order to be sensitive to any artifacts at the grid boundary.  The remaining panels show that the trailing oscillations continue, but do not generate any additional noise. We note that the end result is similar to the case without any refined region.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/WavePool.eps}
\caption{This plot shows, for each column, three snapshots at $t=0.2, 0.3$ and 0.8 of a linear wave as it propagates through the domain.  A static refined region extends from $x = 0.25$ to 0.75 and is shown in grey in each panel.  The individual cells are also color-coded by level: blue indicates the root grid, and red is for the refined region.  Each row shows the result for a different hydrodynamic solver (top is PPM, middle is Zeus, and the bottom is MUSCL).   The solid line shows the analytic solution for a linear, undamped wave.  Note that we focus each panel on a small region of the entire domain to better show the wave itself.}
\label{fig.wavepool}
\end{center}
\end{figure}

>>>>

Figure~\ref{fig.wavepool} shows the evolution of the wave at three different times ($t = 0.2, 0.3$ and 0.8) for three of our solvers (PPM, \zeus, and MUSCL).  In all panels, the shaded regions denote the statically refined region. The leftmost column shows the wave just before it enters the refined region so that we can gauge how the solver is operating in the absence of AMR.  The center column shows the wave after it has fully entered the refined region.  The rightmost column shows the wave well after it has exited the refined region.

@@

For the \zeus\ solver, we see that even before it has entered the refined region there are small oscillations excited behind the wave, although it is also worth noting that the wave itself is beautifully propagated without significant damping or phase errors.  In this test we use only our standard, low amount of quadratic artificial viscosity.  These waves could be damped by additional viscosity, but we do not add any in order to be sensitive to any artifacts at the grid boundary.  The remaining panels show that the trailing oscillations continue, but do not generate any additional noise. We note that the end result is similar to the case without any refined region.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/WavePool.eps}
\caption{This plot shows, for each column, three snapshots at $t=0.2, 0.3$ and 0.8 of a linear wave as it propagates through the domain.  A static refined region extends from $x = 0.25$ to 0.75 and is shown in grey in each panel.  The individual cells are also color-coded by level: blue indicates the root grid, and red is for the refined region.  Each row shows the result for a different hydrodynamic solver (top is PPM, middle is \zeus, and the bottom is MUSCL).   The solid line shows the analytic solution for a linear, undamped wave.  Note that we focus each panel on a small region of the entire domain to better show the wave itself.}
\label{fig.wavepool}
\end{center}
\end{figure}
